Reproducible Summary Tables with the \pkg{gtsummary} Package

### 2021-05-18 gtsummary_submission.tex
overleaf

To showcase \pkg{gtsummary} functions, we will use a simulated clinical trial data set containing baseline characteristics of 200 patients who received Drug A or Drug B, as well as the outcomes of tumor response and death.
Each variable in the data frame has been assigned an attribute label with the \pkg{labelled} package \citep{labelled}, e.g. \code{trial \%>\% set\_variable\_labels(age = "Age")} that will be shown in the summary tables. 
These labels are displayed in the \pkg{gtsummary} tables by default, and had labels not been assigned, the variable name would have been shown.

@@

The default output from \code{tbl\_summary()} is meant to be publication ready.
The \code{tbl\_summary()} function can take, at minimum, a data frame as the only input, and returns descriptive statistics for each column in the data frame.
This is often the first table of clinical manuscripts and describes characteristics of the study cohort.
A simple example is shown below.
Notably, by specifying the \code{by=} argument, you can stratify the summary table. 
In the example below, we have split the table by the treatment a patient received.

@@

The function is highly customizable, and it is initiated with sensible default settings.
Specifically, \code{tbl\_summary} detects variable types of input data and calculates descriptive statistics accordingly.
For example, variables coded as \code{0/1}, \code{TRUE/FALSE}, and \code{Yes/No} are presented dichotomously.
Additionally, \code{NA} values are recognized as missing and listed as unknown, and if a data set is labelled, the label attributes are automatically utilized.

@@

In the example below, the \code{"age"} variable's label is updated to \code{"Patient Age"}.
Default summary statistics for both continuous and categorical variables are updated using the \code{statistic=} argument. 
\pkg{gtsummary} uses \CRANpkg{glue} \citep{glue} syntax construct the statistics displayed in the table.
Function names appearing in curly brackets will be replaced by the evaluated value.
The \code{digits=} argument is used to increase the number of decimal places to which the statistics are rounded, and the missing row is omitted with \code{missing = "no"}.

@@

The \code{tbl\_svysummary()} function is analagous to \code{tbl\_summary()} except an \CRANpkg{survey} \citep{survey} object is supplied rather than a data frame.
The summary statistics presented take into account the survey weights, as do any p-values presented.

@@

The \pkg{gtsummary} package utilizes the \pkg{gt} package to print the summary tables.
The \pkg{gt} package exports approximately one hundred functions to customize and style tables.
When you need to add additional details or styling not available within \pkg{gtsummary}, use the \code{as\_gt()} function to convert the \pkg{gtsummary} object to gt and continue customization.

@@

In the example below, the logistic regression model is summarized with \code{tbl\_regression()}.
Note that a reference row for Grade has been added, and the variable labels have been carried through into the table.
Using \code{exponentiate = TRUE}  we exponentiate the regression coefficients, yielding the odds ratios.
The helper function \code{add\_global\_p()} was used to replace the p-values for each term with the global p-value for grade.

@@

The \code{tbl\_regression()} function leverages the huge effort behind the \pkg{broom} package's \code{tidy()} function to perform the initial formatting of the regression object \citep{broom}.
Because \code{tbl\_regression()} utilizes \code{broom::tidy()}, there are many model types that are supported out of the box, such as \code{lm()}, \code{glm()}, \code{lme4::lmer()}, \code{lme4::glmer()}, \code{geepack::geeglm()}, \code{survival::coxph()}, \code{survival::survreg()}, \code{survival::clogit()}, \code{nnet::multinom()}, \code{rstanarm::stan\_glm()}, models built with the \pkg{mice} package \citep{mice} and many more. 
A custom tidier may be specified as well, which is helpful when you need to present non-standard modifications to your model results such as Wald confidence intervals, or results with modified variance-covariance regression coefficients.

@@

The \pkg{gtsummary} tables shown above are often ready for publication as they are; however, it is common that more complex tables need to be constructed. This can be achieved by merging or stacking \pkg{gtsummary} tables using the \code{tbl\_merge()} and \code{tbl\_stack()} functions.
For example in cancer research, we often report models predicting a tumor's response to treatment and risk of death.
These two models are often reported side-by-side, and it's simple to construct using \code{tbl\_merge()}.
First, build a table for each regression model using \code{tbl\_regression()}, then merge the two tables with \code{tbl\_merge()}.
Any number of \pkg{gtsummary} tables can be merged with this function.

@@

Tables printed with \pkg{gtsummary} can be seamlessly integrated into R markdown documents and knitted into various output types using a number of print engines.
The package was written to be a companion to the \pkg{gt} package from RStudio and is optimized to leverage the advanced customization features of this print engine, but offers compatibility with a variety of popular printing methods including \code{knitr::kable()} \citep{knitr}, \CRANpkg{flextable}  \citep{flextable}, \CRANpkg{huxtable} \citep{huxtable}, and \CRANpkg{kableExtra} \citep{kableExtra}.
While \pkg{gt} is used as default for most outputs, you can easily use your print engine of choice with the conversion helper functions provided in the package (e.g. \code{as\_flex\_table()}).
It's possible to get results in HTML, PDF (via \LaTeX), RTF, Microsoft Word, PowerPoint, Excel and others, utilizing the various print engines.
The package is designed to interact with these print engines behind the scenes to reduce burden on users, and you generally only need to be aware of them if you want to add advanced customizations.

>>>>

To showcase \pkg{gtsummary} functions, we will use a simulated clinical trial data set containing baseline characteristics of 200 patients who received Drug A or Drug B, as well as the outcomes of tumor response and death.
Each variable in the data frame has been assigned an attribute label with the \pkg{labelled} package \citep{labelled}, e.g. \code{trial \%>\% set\_variable\_labels(age = "Age")}, that will be shown in the summary tables. 
These labels are displayed in the \pkg{gtsummary} tables by default, and had labels not been assigned, the variable name would have been shown.

@@

The default output from \code{tbl\_summary()} is meant to be publication-ready.
The \code{tbl\_summary()} function can take, at minimum, a data frame as the only input, and returns descriptive statistics for each column in the data frame.
This is often the first table of clinical manuscripts and describes characteristics of the study cohort.
A simple example is shown below.
Notably, by specifying the \code{by=} argument, you can stratify the summary table. 
In the example below, we have split the table by the treatment a patient received.

@@

The function is highly customizable, and it is initiated with sensible default settings.
Specifically, \code{tbl\_summary} detects variable types of input data and calculates descriptive statistics accordingly.
For example, variables coded as \code{0/1}, \code{TRUE/FALSE}, and \code{Yes/No} are presented dichotomously.
Additionally, \code{NA} values are recognized as missing and listed as unknown, and if a data set is labelled, the label attributes are utilized.

@@

In the example below, the \code{"age"} variable's label is updated to \code{"Patient Age"}.
Default summary statistics for both continuous and categorical variables are updated using the \code{statistic=} argument. 
\pkg{gtsummary} uses \CRANpkg{glue} \citep{glue} syntax to construct the statistics displayed in the table.
Function names appearing in curly brackets will be replaced by the evaluated value.
The \code{digits=} argument is used to increase the number of decimal places to which the statistics are rounded, and the missing row is omitted with \code{missing = "no"}.

@@

The \code{tbl\_svysummary()} function is analogous to \code{tbl\_summary()} except a \CRANpkg{survey} \citep{survey} object is supplied rather than a data frame.
The summary statistics presented take into account the survey weights, as do any p-values presented.

@@

The \pkg{gtsummary} package utilizes the \pkg{gt} package to print the summary tables.
The \pkg{gt} package exports approximately one hundred functions to customize and style tables.
When you need to add additional details or styling not available within \pkg{gtsummary}, use the \code{as\_gt()} function to convert the \pkg{gtsummary} object to \pkg{gt} and continue customization.

@@

In the example below, the logistic regression model is summarized with \code{tbl\_regression()}.
Note that a reference row for Grade has been added, and the variable labels have been carried through into the table.
Using \code{exponentiate = TRUE}, we exponentiate the regression coefficients, yielding the odds ratios.
The helper function \code{add\_global\_p()} was used to replace the p-values for each term with the global p-value for grade.

@@

The \code{tbl\_regression()} function leverages the huge effort behind the \pkg{broom} package's \code{tidy()} function to perform the initial formatting of the regression object \citep{broom}.
Because \code{tbl\_regression()} utilizes \code{broom::tidy()}, there are many model types that are supported out of the box, such as \code{lm()}, \code{glm()}, \code{lme4::lmer()}, \code{lme4::glmer()}, \code{geepack::geeglm()}, \code{survival::coxph()}, \code{survival::survreg()}, \code{survival::clogit()}, \code{nnet::multinom()}, \code{rstanarm::stan\_glm()}, models built with the \pkg{mice} package \citep{mice} and many more. 
A custom tidier may be specified as well, which is helpful when you need to present non-standard modifications to your model results such as Wald confidence intervals, or results with modified variance-covariance standard errors.

@@

The \pkg{gtsummary} tables shown above are often ready for publication as they are; however, it is common that more complex tables need to be constructed. This can be achieved by merging or stacking \pkg{gtsummary} tables using the \code{tbl\_merge()} and \code{tbl\_stack()} functions.
For example, in cancer research we often report models predicting a tumor's response to treatment and risk of death.
These two models are often reported side-by-side, and it's simple to construct using \code{tbl\_merge()}.
First, build a table for each regression model using \code{tbl\_regression()}, then merge the two tables with \code{tbl\_merge()}.
Any number of \pkg{gtsummary} tables can be merged with this function.

@@

Tables printed with \pkg{gtsummary} can be seamlessly integrated into RMarkdown documents and knitted into various output types using a number of print engines.
The package was written to be a companion to the \pkg{gt} package from RStudio and is optimized to leverage the advanced customization features of this print engine, but offers compatibility with a variety of popular printing methods including \code{knitr::kable()} \citep{knitr}, \CRANpkg{flextable}  \citep{flextable}, \CRANpkg{huxtable} \citep{huxtable}, and \CRANpkg{kableExtra} \citep{kableExtra}.
While \pkg{gt} is used as default for most outputs, you can easily use your print engine of choice with the conversion helper functions provided in the package (e.g. \code{as\_flex\_table()}).
It's possible to get results in HTML, PDF (via \LaTeX), RTF, Microsoft Word, PowerPoint, Excel and others, utilizing the various print engines.
The package is designed to interact with these print engines behind the scenes to reduce burden on users, and you generally only need to be aware of them if you want to add advanced customizations.
